% !TEX root = main.tex
%----------------------------------------------------------------------
\chapter{Introduction}\label{chap:intro}
%----------------------------------------------------------------------

%----------------------------------------------------------------------
\section{Timetable}
%----------------------------------------------------------------------

In the autumn semester, we will look at the foundations of \textbf{probability theory}. 

\begin{table}[h]
\centering
\begin{tabular}{|llll|}\hline
Day			& Time			& Location	& \\
Mondays		& 11.10 - 12.00	& E/0.15 	& \\
Wednesdays	& 11.10 - 12.00	& Tower 0.02	& \\
Fridays		& 10.00 - 10.50	& E/0.15		& (even weeks only) \\ \hline
\end{tabular}
\end{table}


The material is divided into five blocks of four lectures:

\begin{tabular}{ll}
Weeks 1--2:	& Probability spaces \\
Weeks 3--4:	& Random variables and distributions \\
Weeks 5--6:	& Expectation \\
Weeks 7--8:	& Limit theorems \\ 
Weeks 9--10:	& Joint distributions
\end{tabular}


%----------------------------------------------------------------------
\section{Flipped classroom}
%----------------------------------------------------------------------
The module will run using a \textbf{flipped classroom} approach:
\bit
\it A traditional lecture is primarily a means of transferring information, from the lecturer to the student. The student is expected to assimilate the information, by reading notes and completing exercises, to gain an understanding of the material, ideally before the next lecture.
\it
In a flipped classroom, students are expected to acquire the relevant information \emph{before} attending the class. Contact hours are dedicated to discussion and exercises, leading to assimilation and understanding.
\eit

Reading material will be provided. This will consist of
\ben
\it definitions,
\it theorems and proofs,
\it illustrative examples, and
\it exercises.
\een

\newpage
Before each class, you are expected to
\ben
\it learn the definitions,
\it understand the statements of the theorems,
\it look at the proofs,
\it try some of the exercises.
\een

During each class, we will
\ben
\it clarify definitions, theorems and proofs (where necessary),
\it work on exercises, and
\it assess each other's work.
\een

Points to note:
\bit
\it Printed ``sample answers'' to exercises will \textbf{not} be provided.
\it Answers will be provided via \textbf{screencast}, but strictly on request.
\eit

%----------------------------------------------------------------------
\section{Assessment}
%----------------------------------------------------------------------

\subsection{Formative assessment}
Answers to exercises can be submitted at any time for assessment and feedback.

\subsection{Summative assessment}
Coursework accounts for 20\% of the total marks for the module.
\bit
\it There will be 10 homework assignments, each worth 2\%.
\it Hand out:	Monday of weeks 2/4/6/8/10 (autumn/spring)
\it Hand in:		Friday of weeks 2/4/6/8/10 (autumn/spring)
\eit

\subsection{Assessment criteria}

\fbox{\begin{minipage}{\linewidth}
\centering\vspace*{1ex}
Submitted work should be \textbf{clear}, \textbf{concise} and \textbf{correct}.
\vspace*{1ex}
\end{minipage}}

\vspace*{2ex}
All work will be assessed according to each of these categories, on a scale of 1 to 5.
\vspace*{2ex}

\begin{table}[h]
\centering
\begin{tabular}{| p{2cm} || p{2.0cm} | p{2.0cm} | p{2.0cm} | p{2.0cm} | p{2.0cm} |}\hline
Category		& 1 & 2 & 3 & 4 & 5\\ \hline\hline
\textbf{Clear} & Obscure & Opaque & Fair & Clear & Crystal \\ \hline
\textbf{Concise} & Long-winded or complete\par lack of detail & Too long or\par lack of detail & Fair & Good & Perfect \\ \hline
\textbf{Correct} & Completely incorrect & Major\par mistakes & Fair & Minor\par mistakes & Completely correct \\ \hline
\end{tabular}
\caption*{Draft Assessment Criteria for MA2500}
\end{table}

Formative and summative work should be submitted using the pro-forma shown on the following page.


%----------------------------------------------------------------------
\endinput
%----------------------------------------------------------------------
