% !TEX root = main.tex
%----------------------------------------------------------------------
\begin{exercise}
\begin{questions}
%----------------------------------------
%==========================================================================
\question
Let $X$ and $Y$ be two random variables having the same distribution but which are not necessarily independent. Show that
\[
\cov(X+Y,X-Y)=0
\]
provided that their common distribution has finite mean and variance.

\begin{answer}
Perhaps the simplest method is the following: let $U=X+Y$ and $V=X-Y$. Then 
\begin{align*}
\cov(X+Y,X-Y) 
	& = \expe(UV) - \expe(U)\expe(V) \\
	& = \expe\big[(X+Y)(X-Y)\big] - \expe(X+Y)\expe(X-Y) \\
	& = \expe(X^2 - Y^2) - \big[\expe(X)+\expe(Y)\big]\big[\expe(X)-\expe(Y)\big]	\qquad\text{(by the linearity of expectation)} \\
	& = \expe(X^2) - \expe(Y^2) - \expe(X)^2 +\expe(X)\expe(Y) - \expe(Y)\expe(X) + \expe(Y)^2 \quad\text{(by linearity again)} \\
	& = \big[\expe(X^2) - \expe(X)^2\big] - \big[\expe(Y^2) - \expe(Y)^2\big] \\
	& = \var(X) - \var(Y).
\end{align*}
Since $X$ and $Y$ have the same distribution, their variances must be equal, so $\cov(X+Y,X-Y)=0$.
\end{answer}



%==========================================================================
\question
Consider a fair six-sided die whose faces show the numbers $-2,0,0,1,3,4$. The die is independently rolled four times. Let $X$ be the average of the four numbers that appear, and let $Y$ be the product of these four numbers. Compute $\expe(X)$, $\expe(X^2)$, $\expe(Y)$ and $\cov(X,Y)$.
\begin{answer}
Let $X_1,X_2,X_3,X_4$ be independent discrete random variables on the set $\{-2,0,0,1,3,4\}$. Each $X_i$ is identically distributed according to the following PMF:
\[\begin{array}{|c|ccccc|}\hline
k			& -2		& 0		& 1		& 3		& 4 	\\ \hline
\prob(X_i=k)		& 1/6	& 1/3	& 1/6	& 1/6 	& 1/6	\\ \hline
\end{array}\]
Hence for $i=1,2,3,4$,
\begin{align*}
\expe(X_i)		& = \frac{1}{6}(-2+0+0+1+3+4) = 1, \\
\expe(X_i^2)	& = \frac{1}{6}(4+0+0+1+9+16) = \frac{30}{6} = 5, \\
\var(X_i)		& = \expe(X_i^2)-\expe(X_i)^2 = 4.
\end{align*}

Let $X=\frac{1}{4}(X_1+X_2+X_3+X_4)$. Then
\[
\expe(X) 	= \frac{1}{4}\big(\expe(X_1)+\ldots+\expe(X_4)\big) = 1
\]
By independence,
\[
\var(X) 	= \frac{1}{16}\big(\var(X_1)+\ldots+\var(X_4)\big) = 1 
\]
so $\expe(X^2)	= \var(X) + \expe(X)^2 = 2$.

and $Y=X_1 X_2 X_3 X_4$. By independence,
\begin{align*}
\expe(Y) 
	& = \expe(X_1 X_2 X_3 X_4) \\
	& = \expe(X_1)\expe(X_2)\expe(X_3)\expe(X_4) \\
	& = 1 \\
\end{align*}
and because the $X_i$ are identically distributed,
\begin{align*}
\expe(XY) 
	& = \frac{1}{4}\expe\big(\expe(X_1+X_2+X_3+X_4)X_1X_2X_3X_4\big) \\
	& = \expe(X_1^2)\expe(X_2)\expe(X_3)\expe(X_4) \\
	& = \expe(X_1^2) = 5 \\
\end{align*}
so $\cov(XY) = \expe(XY) - \expe(X)\expe(Y) = 4$.
\end{answer} 

%==========================================================================
\question
A fair die is rolled twice. Let $U$ denote the number obtained on the first roll, let $V$ denote the number obtained on the second roll, let $X=U+V$ denote their sum and let $Y=U-V$ denote their difference. Compute the mean and variance of $X$ and $Y$, and compute $\expe(XY)$. Check whether $X$ and $Y$ are uncorrelated. Check whether $X$ and $Y$ are independent.
\begin{answer}
Let $U,V\sim\text{Uniform}\{1,2,3,4,5,6\}$ be independent (and identically distributed) random variables, and define $X=U+V$ and $Y=U-V$. 
\begin{align*}
\expe(X)	& = \expe(U) + \expe(V) = 7 \\
\expe(Y)	& = \expe(U) - \expe(V) = 0 \\
\intertext{By independence,}
\var(X)		& = \var(U) + \var(V) = 35/6 \\
\var(Y)		& = \var(U) + \var(V) = 35/6 \\
\intertext{Because $U$ and $V$ are identically distributed, and}
XY 			& = (U+V)(U-V) = U^2 - V^2 \\
\intertext{it follows that}
\expe(XY)	& = \expe(U^2) - \expe(V^2) = 0 \\
\intertext{$X$ and $Y$ are uncorrelated, since}
\cov(X,Y)	& = \expe(XY) - \expe(X)\expe(Y) =0 \\
\intertext{However $X$ and $Y$ are not independent, because (for example)}
\prob(Y=0)		& \neq \prob(Y=0|X=12)=1
\end{align*}
\end{answer} 



%----------------------------------------
\end{questions}
\end{exercise}
%----------------------------------------------------------------------

%======================================================================
\endinput
%======================================================================
