% !TEX root = main.tex
%======================================================================
\chapter{Classical Probability}\label{chap:classical}
%======================================================================

%----------------------------------------------------------------------
\section{The principle of indifference}
%----------------------------------------------------------------------
Let $\Omega$ be a finite sample space. If the outcomes are indistinguishable (except for their names), the \emph{principle of indifference} states that each outcome should be assigned equal probability.
\bit
%\it The probability of every outcome is $1/n$.
\it We say that the outcomes are \emph{equally likely} to occur.
\it This is often an implicit assumption (e.g. ``a card is chosen at random'').
\eit

The probability of an event $A\subseteq\Omega$ is proportional to its \emph{cardinality}:
\[
\prob(A) = \displaystyle\frac{|A|}{|\Omega|}.
\]

\bit
\it Classical examples involve \emph{coins}, \emph{dice} and \emph{cards}.
\it Problems can be solved by counting the number of ways that different events can occur.
\eit

%----------------------------------------------------------------------
\section{Sampling with replacement}
%----------------------------------------------------------------------
In many cases, once an element has been chosen it is \emph{replaced} in the set, and can be chosen again in subsequent selections. If there are $n$ possible elements, there are $n^k$ distinct choices of $k$ elements under this selection model.

% example: choosing with replacement
\begin{example} 
There are exactly $2^k$ binary sequences of length $k$, $3^k$ ternary sequences of length $k$, etc.
\end{example}

% example: choosing with replacement
\begin{example}
UK registration plates are of the form \texttt{\,AB\,12\,CDE\,}: the first two positions can be any letters except \texttt{I}, \texttt{Q} and \texttt{Z}, the third and fourth positions can be any single digit, and the last three positions can be any letters except \texttt{I} and \texttt{Q}. 
How many different UK vehicle registration plates are possible?
\end{example}
\begin{solution}
The total number of possible registration plates is 
\[
23 \times 23 \times 10 \times 10 \times 24 \times 24 \times 24 = 731\,289\,600.
\]
\end{solution}

\newpage

%----------------------------------------------------------------------
\section{Sampling without replacement}
%----------------------------------------------------------------------
It also happens that once an element has been chosen, it cannot be chosen again in subsequent selections. There are two cases here, depending on whether or not we take the \emph{order} in which elements are chosen into account.

\begin{definition}
\ben
\it A \emph{$k$-permutation} of a set of $n$ elements is an ordered sequence of $k$ elements, taken (without replacement) from the set. The number of $k$-permutations is
\[
^nP_k = n(n-1)(n-2)\cdots(n-k+1) = \frac{n!}{(n-k)!}
\]
\it A \emph{$k$-combination} of a set of $n$ elements is an un-ordered subset of $k$ elements, taken (without replacement) from the set. The number of $k$-combinations is
\[
^nC_k = \frac{n(n-1)(n-2)\cdots(n-k+1)}{k(k-1)(k-2)\cdots 1} = \frac{n!}{(n-k)!k!}
\]
This is called the \emph{binomial coefficient}, and is usually denoted by $\binom{n}{k}$.
\een
\end{definition}

% urn model
\begin{example}
An urn contains $3$ white balls and $5$ black balls. Two balls are drawn at random from the urn. What is the probability that they are both white?
\end{example}

\begin{solution}
\bit
\it Let $\Omega$ be the set of all possible pairs: $|\Omega|	= \displaystyle\binom{8}{2} = \frac{8!}{6!2!} = 28$. 
\it Let $A$ be the event that both balls are white: $|A|  = \displaystyle\binom{3}{2} = \frac{3!}{1!2!} = 3$.
\eit
Hence, $\prob(A) = |A|/|\Omega| = 3/28$.
\end{solution}


% example
\begin{example}
Find the probability that a hand of five cards contains (a) four cards of the same kind, and (b) two distinct pairs.
%\ben 
%\it four cards of the same kind (e.g. four aces),
%\it two (distinct) pairs.
%\een
\end{example}

\begin{solution}
Let $\Omega$ be the set of all possible hands: $|\Omega| = \binom{52}{5} = 2598960$.
\ben
\it Let $A$ be the event that the hand contains four of a kind.
	\bit
	\it There are 13 different kinds of card (A,2,3,4,5,6,7,8,9,10,J,Q,K).
	\it For each kind, there are $48$ different hands that contain all four cards of that kind.
	\it Hence there are $|A| = 13\times 48 = 624$ hands containing four cards of the same kind.
	\it Thus $\prob(\text{four-of-a-kind}) = |A|/|\Omega| = 624/2598960 = 0.00024$.
	\eit
\it Let $B$ be the event that the hand contains two pairs.
	\bit
	\it Two (distinct) pairs can be chosen in $\binom{13}{2} = 78$ different ways.
	\it Each pair can be chosen in $\binom{4}{2} = 6$ different ways. 
	\it The fifth card can be chosen in $44$ different ways.
	\it Hence there are $|B| = 78\times 6\times 6\times 44 = 123552$ hands containing two pairs.
	\it Thus $\prob(\text{two-pairs}) = |B|/|\Omega| = 123552/2598960 = 0.0475$.
	\eit
\een	
\end{solution}

% example: division paradox
\begin{example}[The Division Paradox]
A \emph{fair game} is a game in which the probability of winning is equal to the probability of losing. Two players $A$ and $B$ decide to play a sequence of fair games until one of the players wins 6 games, but they stop when the score is 5:3 in favour of player $A$. How should the prize money be fairly divided?
\end{example}

\begin{solution}
Assume that the players carried on playing the sequence of games,
\bit
\it The maximum number of additional games is 3.
\it The possible outcomes are 
\[
\Omega=\{AAA,AAB,ABA,BAA,ABB,BAB,BBA,BBB\}.
\]
\eit
The games are fair, so all outcomes are equally likely.
\bit
\it Only one outcome is in favour of player $B$, the other seven are in favour of $A$.
\it The prize money should therefore be divided in the ratio $7:1$ in favour of $A$.
\eit
\end{solution}

% example: birthday paradox
\begin{example}[The Birthday Paradox]
Find the smallest number of people for which the probability that least two share a birthday exceeds $1/2$.
\end{example}

\begin{solution}
\bit
\it Suppose we have a set of $k$ randomly chosen people.
\it Let $n=365$ be the number of days in a year. 
\it Assume that each birthday is equally likely to occur.
\eit
Let $p(n,k)$ be the probability that at least two people share a birthday.
\bit
\it We want to find the value of $k$ for which $p(n,k) > 0.5$.
\it There are $n$ choices of birthday for each of the $k$ people, so the total number of outcomes is $n^k$. 
\it The number of outcomes in which all birthdays are different is the number of $k$-permutations of $n$ elements, which is equal to 
\[
^nP_k = n(n-1)(n-2)\cdots(n-k+1) = \frac{n!}{(n-k)!}
\]
\eit
Hence,
\begin{align*}
p(n,k)
	& = \prob(\text{At least one pair share a birthday}) \\
	& =  1 - \prob(\text{No pairs share a birthday}) \\
	& =  1 - \frac{\text{Number of outcomes with all birthdays different}}{\text{Total number of outcomes}} \\
	& =  1 - \frac{n!}{n^k(n-k)!}
\end{align*}
The probabilities $p(365,k)$ for various values of $k$ are shown below:
\[
\begin{array}{|l|cccccccc|} \hline
k			& 2		& 10		& 15		& 22		& 23		& 50 	& 100		& 366 	\\ \hline
p(k,365)		& 0.003	& 0.12	& 0.25	& 0.48	& 0.51	& 0.97	& 0.99997 & 1		\\ \hline
\end{array}
\]
\end{solution}

%----------------------------------------------------------------------
\section{Exercises}
\input{ex06_classical}
%----------------------------------------------------------------------

%======================================================================
\endinput
%======================================================================
